---
title: "Tutorial to use scPipeVis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction-to-scPipeVis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(scPipeVis)
```

## Introduction

Single cell RNA sequencing (scRNA-seq) is a genomic approach used to quantify messenger RNA sequences (whole transcriptome) at the resolution of a single cell, which can be scaled to tens of thousands of cells at the same time. This results in generation of a cell by gene counts matrix (in R gene by cell) as the primary data, and additional metadata from any given scRNA-seq experiment. Data transformation, modelling, and analysis methods operate downstream on the above described data. `scPipeVis` is an R package that lets you perform all the steps with a single function so users can download the package and run this function to do a very preliminary bare-bones data analysis of scRNA-seq data. Additionally there is a visualisation function that creates basic plots which are typically used for the visualisation of scRNA-seq data. The complete motivation behind the creation of this package follows. There are major limitations of this package which are also talked about below.

## Motivation

A default single cell RNA sequencing data analysis pipeline comprises of several steps, from quality control, normalization of count matrix, feature selection, dimensionality reduction, clustering, annotating cell types, and then differential gene expression followed by gene set enrichment analysis. In my literature review, I found that all of these steps can be performed in so many different ways by using different methods at each step. This process is further complicated by tuning of hyperparameters at many different steps. Therefore, any user while analyzing their data, has to make so many choices/decisions at each step. These decisions, like tuning the hyperparameter or choosing the right method is absolutely not trivial. In other words, these choices at each step can affect your downstream analyses. Therefore, making well informed decisions are crucial. This is where scPipeVis comes in, following an extensive literature review, I have put together a pipeline that takes care of most of these above mentioned steps using the current best practices methods based on published reviews benchmarking different methods for all these steps (Amezquita et al. 2019). Further scPipeVis provides a function to visualize each step of the pipeline with many different and appropriate depictions. These depictions are chosen after going through the literature and figuring out what kind of visualizations are required at each step. Therefore, this part of the package will facilitate informed decision making for users as they go on to select the best suitable methods and hyper parameters which in turn best suits their own datasets and meets their needs.

## Limitations of `scPipeVis`

This package is created for preliminary data analysis of scRNA-seq data, at the end of the day the user will have to eventually figure out if certain methods work better to answer the question that they are trying to answer using their data. The user will also eventually have to tune hyperparameters of certain functions, if they dont get satisfactory results from this pipeline. Last but not the least, here I don't provide an exhaustive set of visualizations. I will continue working on the package to add more and more visualizations and make the pipeline more end to end.

## What the package provides

`scPipeVis` includes four main components:

1.  `run_pipeline()` Runs a standard scRNA-seq preprocessing workflow: QC summarization, log-normalization, highly variable gene detection, PCA, UMAP, and unsupervised clustering.

2.  `visualise_pipeline()` Generates diagnostic plots including the HVG mean–variance trend, PCA and UMAP embeddings, and simple QC/clustering summaries.

3.  `plot_gene_on_umap()` Produces a viridis-coloured UMAP plot showing the expression of a user-specified gene.

4.  `run_scPipeVis_app()` Launches the Shiny application for interactive gene-on-UMAP visualization. The app can use either a built-in demo dataset or a user-provided `SingleCellExperiment`.

## Loading and preparing data

We will demonstrate the workflow using a public dataset from the `scRNAseq` package. If needed, install it using:

```{r}
# install.packages("BiocManager")
BiocManager::install("scRNAseq")
```

Load the dataset (Zeisel et al., 2015 mouse brain) and inspect the object:

```{r}
library(scRNAseq)
library(SingleCellExperiment)

sce <- ZeiselBrainData()
sce
```

This dataset contains mouse brain cells with raw counts and cell metadata in a `SingleCellExperiment`, which is the input format expected by `scPipeVis`

## Running the pipeline

`run_pipeline()` performs the following steps:

-   **QC summarization** using `scater::addPerCellQC()`

-   **Log-normalization** using `scuttle::logNormCounts()`

-   **HVG detection** using `scran::modelGeneVar()` and `scran::getTopHVGs()`

-   **PCA** using HVGs via `scran::fixedPCA()`

-   **UMAP embedding** via `scater::runUMAP()`

-   **Graph-based clustering** using an SNN graph (`scran::buildSNNGraph()`)

Run the pipeline:

```{r}
library(scPipeVis)
sce_processed <- run_pipeline(sce)
sce_processed
```

Inspect available reduced dimensions:

```{r}
names(reducedDims(sce_processed))
```

InExamine HVG model results:

```{r}
metadata(sce_processed)$hvg_data
```

Check clustering:

```{r}
table(colData(sce_processed)$cluster_id)

```

## Visualizing workflow results

`visualise_pipeline()` produces plots illustrating major steps:

```{r}
visualise_pipeline(sce_processed)
```

This includes:

-   the mean–variance trend from HVG selection

-   a PCA embedding

-   a UMAP embedding

-   automatic colouring based on `cluster_id`

These plots are useful for diagnosing whether the structure of the data is captured appropriately.

## Gene expression visualisation

To explore how individual genes behave across clusters in the UMAP embedding, use:

```{r}
plot_gene_on_umap(sce_processed, gene = "Gad1")

```

This enables identification of cluster-specific markers.

## Saving and reloading results

`SingleCellExperiment` objects can be saved and restored easily:

```{r}
# Uncomment to run
# saveRDS(sce_processed, file = "sce_processed.rds")
# sce_reloaded <- readRDS("sce_processed.rds")
```

## Interactive Shiny App

`scPipeVis` includes an interactive Shiny application designed for quick exploratory analysis without writing code.

Launch the app:

```{r}
# Uncomment to run
# run_scPipeVis_app()

```

or supply your own dataset:

```{r}
# Uncomment to run
# run_scPipeVis_app(sce = sce_processed)

```

The Shiny app:

-   runs the pipeline internally

-   provides a searchable dropdown of all genes

-   displays a clean UMAP plot of the selected gene's expression

-   includes summary statistics for the chosen gene

## More thing to add

-   Automated annotations

## References

-   Amezquita RA, Lun ATL, Becht E, Carey VJ, Carpp LN, Geistlinger L, Marini F, Rue-Albrecht K, Risso D, Soneson C, et al. 2019 Dec 2. Orchestrating single-cell analysis with Bioconductor. Nature Methods. <doi:https://doi.org/10.1038/s41592-019-0654-x>.
-   Hadley Wickham. 2016. ggplot2 Elegant Graphics for Data Analysis. Cham Springer International Publishing.
-   Hadley Wickham. 2019. Advanced R. Boca Raton: Chapman & Hall/Crc.
-   Lun ATL, McCarthy DJ, Marioni JC. 2016. A step-by-step workflow for low-level analysis of single-cell RNA-seq data with Bioconductor. F1000Research. 5:2122. <doi:https://doi.org/10.12688/f1000research.9501.2>.
-   McCarthy DJ, Campbell KR, Lun ATL, Wills QF. 2017 Jan 14. Scater: pre-processing, quality control, normalization and visualization of single-cell RNA-seq data in R. Bioinformatics.:btw777. <doi:https://doi.org/10.1093/bioinformatics/btw777>.
-   Risso D, Cole M (2025). scRNAseq: Collection of Public Single-Cell RNA-Seq Datasets. <doi:10.18129/B9.bioc.scRNAseq>, R package version 2.23.1, <https://bioconductor.org/packages/scRNAseq>.
-   Wickham H. 2015. R Packages. “O’Reilly Media, Inc.”
-   Slides for BCB410 by professor Anjali Silva

## Session information {.unnumbered}

```{r}
sessionInfo()
```
